Kindly Bent to Free Us

### 2020-05-30 kindly/motivation.tex
Tiny presentation fix.
Small code fixes for the new iterators.

This approach is not sufficient if we want to mutate the array during iteration.
To this end, we need to take an exclusive borrow of the structure to iterate
on:
\begin{lstlisting}[numbers=none]
val iter_mut : (&!(aff_r1,'a) -{un}> unit) -> &!('k_1,'a Array.t) -> unit
\end{lstlisting}

The callback is explicitly annotated with @un@ because  it will be called multiple times.

@@

To close this discussion, let's see which primitives are needed to implement
functions like @bfold@ and @iter_mut@. The naive sequential implementations of
both functions boil down to a loop over the index range of the array:
\begin{flushleft}\vspace{-\baselineskip}
  \begin{minipage}[t]{0.49\linewidth}
\begin{lstlisting}
let rec bfold_helper f a z i =
  if i < 0 then z
  else bfold_helper f &a (f (get_sb (&a, i))) (i-1)
let bfold f a z =
  bfold_helper f a z (length &a - 1)
\end{lstlisting}
  \end{minipage}
  \begin{minipage}[t]{0.49\linewidth}
\begin{lstlisting}
let rec iter_helper f a i =
  if i < 0 then ()
  else (f (get_eb (&!a, i)); iter_helper f &!a (i-1))
let iter_mut f a =
  iter_mut f a (length &a - 1)
\end{lstlisting}
  \end{minipage}
\end{flushleft}
Observe that we are proposing two different primitives
\begin{itemize}
\item @get_sb : &('k, 'a Array.t) -> int -> &('k, 'a)@ \\to get a
  shared borrow from a shared borrow of an array and
\item @get_eb :  &!('k, 'a Array.t) -> int -> &!('k, 'a)@ \hfill (unsafe!) \\to get an
  exclusive borrow from an exclusive borrow of  an array.
\end{itemize}
These primitives have the same underlying implemention (the same as
@get@ from \cref{sec:imper-progr}). Their types arise from the
intuition that the borrow of a structure should entitle to borrows of
its substructures, roughly, the borrow of an array could be considered as
array of borrows.
However, only @get_sb@ is safe: the shared borrow of the array entitles us to
obtain shared borrows for the elements \emph{in the same region} as
the shared borrow for the array freezes the array inside the
region. This freeze extends to the elements because 
lifetime of the array fully overlaps with the lifetime of its
elements. 
Considering @get_eb@, we see that we may obtain \emph{two different
exclusive borrows} of the same array element inside a region. Clearly,
the exclusive borrow for the element should live in a nested region
where the array is not accessible. Hence, the safe alternative is
to use a different function for obtaining borrows of elements
\begin{lstlisting}[numbers=none]
with_eborrow : ('b:'k),('k <= lin_r) => &!('k_1,'a Array.t) -> int -> (&!(aff_r1,'a) -> 'b) -{'k_1}> 'b
\end{lstlisting}
Hence, the helper function for @iter_mut@ should read like this
\begin{lstlisting}
let rec iter_helper f a i =
  if i < 0 then ()
  else (with_eborrow &!a i f; iter_helper f &!a (i-1))
\end{lstlisting}
which also explains the occurrence of @aff_r1@ in the type of
@iter_mut@.

>>>>

This approach is not sufficient if we want to mutate the elements of
the array during iteration.
To this end, we need to take an exclusive borrow of the structure to iterate
on:
\begin{lstlisting}[numbers=none]
val iter_mut : (&!(aff_r1,'a) -> unit) -> &!('k_1,'a Array.t) -> unit
\end{lstlisting}

@@

To close this discussion, let's see which primitives are needed to implement
functions like @bfold@ and @iter_mut@. The naive sequential implementations of
both functions boil down to a loop over the index range of the array:
\begin{flushleft}\vspace{-\baselineskip}
  \begin{minipage}[t]{0.49\linewidth}
\begin{lstlisting}
let rec bfold_helper f a z i =
  if i < 0 then z
  else bfold_helper f &&a (f (get_sb &&a i) z) (i-1)
let bfold f a z =
  let l = length &&a - 1 in
  bfold_helper f &&a z l
\end{lstlisting}
  \end{minipage}
  \begin{minipage}[t]{0.49\linewidth}
\begin{lstlisting}
let rec iter_helper f a i =
  if i < 0 then ()
  else (f (get_eb &&!a i); iter_helper f &&!a (i-1))
let iter_mut f a =
  let l = length &&a - 1 in
  iter_helper f &&!a l
\end{lstlisting}
  \end{minipage}
\end{flushleft}
Observe that we are proposing two different primitives
\begin{itemize}
\item
  @get_sb : &('k, 'a Array.t) -> int -{'k}> &('k, 'a)@ \\to get a
  shared borrow from a shared borrow of an array and
\item
  @get_eb : &!('k, 'a Array.t) -> int -{'k}> &!('k, 'a) (* Unsafe! *)@
\\to get an
  exclusive borrow from an exclusive borrow of  an array.
\end{itemize}
These primitives have the same underlying implemention (the same as
@get@ from \cref{sec:imper-progr}). Their types arise from the
intuition that the borrow of a structure should entitle to borrows of
its substructures, roughly, the borrow of an array could be considered as
array of borrows.
However, only @get_sb@ is safe: the shared borrow of the array entitles us to
obtain shared borrows for the elements \emph{in the same region} as
the shared borrow for the array freezes the array inside the
region. This freeze extends to the elements because 
lifetime of the array fully overlaps with the lifetime of its
elements. 
Considering @get_eb@, we see that we may obtain \emph{two different
exclusive borrows} of the same array element inside a region. Clearly,
the exclusive borrow for the element should live in a nested region
where the array is not accessible. Hence, the safe alternative is
to use a different function for obtaining borrows of elements
\begin{lstlisting}[numbers=none]
with_eborrow : ('b:'k),('k <= lin_r) => &!('k_1,'a Array.t) -> int -> (&!(aff_r1,'a) -> 'b) -{'k_1}> 'b
\end{lstlisting}
Hence, the helper function for @iter_mut@ should read like this
\begin{lstlisting}
let rec iter_helper f a i =
  if i < 0 then ()
  else (with_eborrow &!a i f; iter_helper f &!a (i-1))
\end{lstlisting}
which also explains the occurrence of @aff_r1@ in the type of
@iter_mut@.
